Chemistry of the Orphan Stream I: Identifying Stream Candidates with Low Resolution Spectroscopy

### 2012-09-12 2012-oss-mrs/Response_GDC_20120912_II.txt

3 lines from the bottom: "discriminate dwarfs from giants"  (i.e. and ==> from)
>> Changed

2 lines from the bottom "spectra" is missing after "arm".
>> Added

2nd paragraph of 4.1.  Think we need to say that "stars" in the Besancon models were selected in the same way as the observed sample, i.e. same colour and magnitude cuts.  (New Fig 1 is fine).
>> Whoops! I had forgotten to do this after Stefan gave me the true selection criteria. I've done it now though and re-plotted the histogram. The kinematic distribution did not change significantly. I've also added the following as the second sentence in 2nd paragraph of 4.1: " We have selected predicted stars from Besan\c{c}on using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations."

>> In light of the new Besancon histogram, there is a hint of Orphan Stream there in kinematics alone, so I have altered the text:
>> "No statistically significant peak representative of the Orphan Stream exists in our sample"
>> to
>> "There is no sharp kinematic peak representative of the Orphan Stream exists in our sample."
>> The statistical significance is probably questionable, i.e. it could be argued either way, so rather than make a bold statement I've erred on the side of caution -- since we recover OSS members quite reliably using a combination of cuts anyways.

### 2012-09-12 2012-oss-mrs/draft-oss-mrs.tex
Fixed text to match for new Vgsr histogram plot
Minor changes after a readthrough

\author{Andrew R. Casey\altaffilmark{1,2}, Stefan C. Keller\altaffilmark{1}, Gary Da Costa\altaffilmark{1}, Elizabeth Maunder\altaffilmark{1}, Anna Frebel\altaffilmark{2}}
\altaffiltext{1}{Research School of Astronomy \& Astrophysics, Australian National University, Mount Stromlo Observatory, via Cotter Rd, Weston, ACT 2611, Australia; \email{acasey@mso.anu.edu.au}}
\altaffiltext{2}{Massachusetts Institute of Technology, Kavli Institute for Astrophysics and Space Research,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}

@@

The Milky Way stellar halo has partly formed through the accretion of satellites which are disrupted by tidal forces as they fall into the Milky Way's potential. Stars which were once gravitationally bound to the satellite are left behind in a stream of stars. Stream kinematics are sensitive to the shape of the galactic potential, allowing us to constrain the Milky Way potential, and reconstruct the formation history of the Milky Way. The level of accreted substructure in the Milky Way has only recently become apparent through multi-band photometric surveys like the Sloan Digital Sky Survey (SDSS). The more prominent of the detectable substructures, like Sagittarius, have been well-studied. One of the more prominent \-- yet less studied \-- substructures is that of the Orphan Stream.

@@

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of seven highly probable Orphan Stream candidates, all worthy of high resolution spectroscopic follow up. In the following section we outline our photometric target selection. In Section \S\ref{sec:observations} we describe the low resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

Given our colour selection we expect to recover giants and dwarf contaminants. The lack of 2MASS photometry for the selected stars means we cannot use the $(J-H, H-K)$ diagram to aid the separation of candidate giants from contaminating dwarfs.

@@

The data were reduced using the standard \textsc{2DFDR} reduction pipeline \footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35\,px$^{-1}$. With the presence of strong Ca \textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg I at 8807\,{\AA}, and the Mg I\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

% Besan\c{c}on heliocentric velocities have been translated to a galactocentric frame in the same way as our targets.
Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from Besan\c{c}on using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. No statistically significant peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In four cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS $g \-- r$ de-reddened\footnote{All magnitudes presented in this paper are de-reddened}, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Given that our four Orphan Stream stars cover range along the giant branch, we are in a good position to revise the distance estimate to the stream. Given a 10 Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.74$, we find a best-fitting distance to the stream of 24 kpc at ($l, $b) = ($250^\circ$, $50^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$ kpc found by \citet{Newberg;et-al_2010}, but given the inherent uncertainties in determining distance we argue these distances are largely in agreement.

@@

If the Orphan Stream continues through SEGUE stripe 1540 at $(l, b) = (271\,^\circ$, $38\,^\circ)$ as \citet{Newberg;et-al_2010} found, then the stream is even closer there than in the region analysed here. Thus, if our observations and analyses are repeated at $(271\,^\circ$, $38\,^\circ)$, we predict K giant stream members of brighter apparent magnitude will be recovered. If the stream density is similar to or higher than what we observe at $(l, b) = (250\,^\circ, 50\,^\circ)$ then more K giant Orphan Stream members are likely to be recovered in the stripe 1540 field.

>>>>

\author{Andrew R. Casey\altaffilmark{1,2}, Stefan C. Keller\altaffilmark{1}, Gary Da Costa\altaffilmark{1}, Elizabeth Maunder\altaffilmark{1}}
\altaffiltext{1}{Research School of Astronomy \& Astrophysics, Australian National University, Mount Stromlo Observatory, via Cotter Rd, Weston, ACT 2611, Australia; \email{acasey@mso.anu.edu.au}}
\altaffiltext{2}{Massachusetts Institute of Technology, Kavli Institute for Astrophysics and Space Research,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}

@@

The Milky Way stellar halo has partly formed through the accretion of satellites which are disrupted by tidal forces as they fall into the Milky Way's potential. Stars which were once gravitationally bound to the satellite are left behind in a stream of stars. Stream kinematics are sensitive to the shape of the galactic potential, allowing us to constrain the Milky Way potential, and reconstruct the formation history of the Milky Way. The level of accreted substructure in the Milky Way has only recently become apparent through multi-band photometric surveys like the Sloan Digital Sky Survey (SDSS). The more prominent of the detectable substructures, like Sagittarius, have been well-studied. One of the more prominent \--- yet less studied \--- substructures is that of the Orphan Stream.

@@

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of X highly probable Orphan Stream candidates, all worthy of high resolution spectroscopic follow up. In the following section we outline our photometric target selection. In section \ref{sec:observations} we describe the low resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

Given our colour selection we expect to recover giants, and dwarf contaminants. The lack of 2MASS photometry for the selected stars means we cannot use the $(J-H, H-K)$ diagram to aid the separation of candidate giants from contaminating dwarfs.

@@

The data were reduced using the standard \textsc{2DFDR} reduction pipeline\footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35\,px$^{-1}$. With the presence of strong Ca \textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg I at 8807\,{\AA}, and the Mg I\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

% Besan\c{c}on heliocentric velocities have been translated to a galactocentric frame in the same way as our targets.
Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from Besan\c{c}on using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no sharp kinematic peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS $g \-- r$ de-reddened\footnote{All magnitudes presented in this letter are de-reddened}, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Given that our four Orphan Stream stars cover range along the giant branch, we are in a good position to revise the distance estimate to the stream. Given a 10 Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.78$, we find a best-fitting distance to the stream of 24 kpc at ($l, $b) = ($250^\circ$, $50^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$ kpc found by \citet{Newberg;et-al_2010}, but given the inherent uncertainties in determining distances we argue these measurements are largely in agreement.

@@

If the Orphan Stream continues through SEGUE Stripe 1540 at $(l, b) = (271\,^\circ$, $38\,^\circ)$ as \citet{Newberg;et-al_2010} found, then the stream is even closer there than in the region analysed here. Thus, if our observations and analyses are repeated at $(271\,^\circ$, $38\,^\circ)$, we predict K giant stream members of brighter apparent magnitude will be recovered. If the stream density is similar to or higher than what we observe at $(l, b) = (250\,^\circ, 50\,^\circ)$ then more K giant Orphan Stream members are likely to be recovered in the Stripe 1540 field.
